---
title: "HW5_PH1700"
author: "Yue Zhang"
date: "2024-10-02"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    keep_tex: true
  bookdown::html_document2:
    df_print: paged
    toc: true
geometry: margin=2.54cm
fontsize: 12pt
mainfont: Times New Roman
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
knitr::opts_knit$set(root.dir = "E:/Biostat/Biostatistics/PHL_1700/")
```

```{r packages, include = FALSE}
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(readxl)
library(haven)
library(stats)

mytheme = theme_clean(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "right",
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
    ),
    plot.title = element_text(size = 12)
  )
theme_set(mytheme)
```

# 8.81
## Research question: Is there a difference in degree of pain during maximal activity while on Motrin compared to placebo?

(a)<br>
```{r 81a, message=FALSE, warning=FALSE, fig.align = "center"}
Tennis = read.csv("/Biostat/Biostatistics/PHL_1700/Data/Raw/Tennis.csv")

#Since we have 88 observations which is larger than 30, we should be able to apply the CLT.

# Create variable to calculate the pain difference (Motrin - Placebo)
Tennis$pain_diff = ifelse(
  Tennis$drg_ord == 1,
  Tennis$painmx_2 - Tennis$painmx_4,
  Tennis$painmx_4 - Tennis$painmx_2
)

# Plot histogram of the differences
hist(
  Tennis$pain_diff,
  main = "Pain Score Difference (Motrin - Placebo)",
  xlab = "Pain Score Difference",
  col = "lightblue",
  breaks = 10
)
```

```{r 81a2, message=FALSE, warning=FALSE, fig.align = "center"}
#Plot boxplot of the difference
boxplot(
  Tennis$pain_diff,
  main = "Pain Score Difference (Motrin - Placebo)",
  xlab = "Pain Score Difference",
  col = "lightblue"
)
```
The histogram shows a reasonably symmetric and bell-shaped distribution, indicating that the normalization of the data. Thus the CLT applies. The boxplot shows that there are very few outliers and the median of pain score difference is around zero, indicating the difference between Motrin and placebo is small. 

(b)<br>
A paired t-test should be used, since this is a cross-over design.

(c)<br>
Null hypothesis: There is no difference in degree of pain during maximal activity while on Motrin compared to placebo
Alternative hypothesis: There is difference in degree of pain during maximal activity while on Motrin compared to placebo

```{r 81c, message=FALSE, warning=FALSE}
#Since we've calculated the pain difference, we'll use one sample t test
t.test(Tennis$pain_diff, mu = 0)
```

The p-value is 0.001 < 0.05, then we have enough evidence to reject the null hypothesis that there is difference in degree of pain during maximal activity while on Motrin compared to placebo. 

(d)<br>
The 95% CI is (0.34, 1.28) which means that we are 95% confident that the difference of the means will lies between 0.34 and 1.28. However, the null value is 0 which is not included in the CI. Thus, we can reject the null hypothesis.

(e)<br>
Null hypothesis: There is no difference in degree of pain during maximal activity while on Motrin compared to placebo
Alternative hypothesis: Motrin is associated with a lower degree of pain during maximal activity compared to placebo
```{r 81e, message=FALSE, warning=FALSE}
t.test(Tennis$pain_diff, mu=0, alternative = "less")
```
The 95% CI is (-inf, 1.20), which means that we are 95% confident that the difference of means will lies in the region. The null value (0) is included in the CI and our p-value (0.9995) is larger than 0.05, thus we don't have enough evidence to reject the null hypothesis.

# 8.139
## Research question: Do boys with better glycemic control have different growth patterns in weight than boys with poorer glycemic control?

(1)<br>
```{r 139a, message=FALSE, warning=FALSE}
Diabetes = read_dta("/Biostat/Biostatistics/PHL_1700/Data/Raw/DIABETES-1.DAT.dta")
write.csv(Diabetes, "diabetes.csv", row.names = FALSE)

#Calculate the average HbgAlc for each subject
Diabetes_clean = Diabetes %>% group_by(id) %>% summarize(avg_hbgAlc = mean(gly_a1c))
head(Diabetes_clean)
```

(2)<br>
```{r 139b, message=FALSE, warning=FALSE}
#Calculate the median HgbAlc for all boys
Median_hbgAlc = median(Diabetes_clean$avg_hbgAlc)
```

(3)<br>
```{r 139c, message=FALSE, warning=FALSE}
#Categorized boys into two groups
Diabetes_clean = Diabetes_clean %>% 
  mutate(group = ifelse(avg_hbgAlc < Median_hbgAlc, "Controlled", "Uncontrolled"))
head(Diabetes_clean)
```

(4)<br> 
```{r 139d, message=FALSE, warning=FALSE}
#Create growth variable
Growth = Diabetes %>% group_by(id) %>% summarize(
  first_weight = first(wt_kg),
  last_weight = last(wt_kg),
  first_age = first(age_yrs),
  last_age = last(age_yrs)
) %>% mutate(growth_rate = (last_weight - first_weight) / (last_age - first_age))

Diabetes_clean = merge(Diabetes_clean, Growth, by = "id")
head(Diabetes_clean)
```

(5)<br>
First, we will check the normality of these two groups
```{r 139e1, message=FALSE, warning=FALSE, fig.align = "center"}
Controlled = Diabetes_clean$growth_rate[Diabetes_clean$group == "Controlled"]
Uncontrolled = Diabetes_clean$growth_rate[Diabetes_clean$group == "Uncontrolled"]
# Plot histogram of growth rate
hist(
  Controlled,
  main = "Growth Rate (Controlled)",
  xlab = "Growth Rate",
  col = "lightblue"
)
```

```{r 139e2, message=FALSE, warning=FALSE, fig.align = "center"}
hist(
  Uncontrolled,
  main = "Growth Rate (Uncontrolled)",
  xlab = "Growth Rate",
  col = "lightblue"
)
```

```{r 139e3, message=FALSE, warning=FALSE, fig.align = "center"}
boxplot(
  Controlled,
  main = "Growth Rate (Controlled)",
  xlab = "Growth Rate",
  col = "lightblue"
)
```

```{r 139e4, message=FALSE, warning=FALSE, fig.align = "center"}
boxplot(
  Uncontrolled,
  main = "Growth Rate (Uncontrolled)",
  xlab = "Growth Rate",
  col = "lightblue"
)
```

```{r 139e5, message=FALSE, warning=FALSE, fig.align = "center"}
qqnorm(Controlled, main = "QQ Plot (Controlled)")
qqline(Controlled, col = "red")
```

```{r 139e6, message=FALSE, warning=FALSE, fig.align = "center"}
qqnorm(Uncontrolled, main = "QQ Plot (Uncontrolled)")
qqline(Uncontrolled, col = "red")
```

```{r 139e7, message=FALSE, warning=FALSE}
shapiro.test(Controlled)
shapiro.test(Uncontrolled)
```

The Shapiro-Wilk test shows that both groups' p-values are less than 0.05, meaning the data significantly deviates from normality. So we cannot test for equal variances. Thus, we'll conduct the two samples t-test with unequal variances

Null hypothesis: There is no difference between the growth rate in controlled group and uncontrolled group
Alternative hypothesis: There is difference between the growth rate in controlled and uncontrolled groups
Alternative hypothesis: The two groups have unequal variances

```{r 8.139e8, message=FALSE, warning=FALSE}
t.test(Controlled, Uncontrolled, alternative = "two.sided")
```

Since the p-value(0.2961) is larger than 0.05, we don't have enough evidence to reject the null hypothesis. Therefore, there is no difference between the growth rate of boys with better glycemic control and boys with pooer glycemic control.

# Additional problems

(a)<br>
```{r a1, message=FALSE, warning=FALSE}
Lead = read_dta("/Biostat/Biostatistics/PHL_1700/Data/Raw/LEAD-1.DAT.dta")
write.csv(Lead, "Lead.csv", row.names = FALSE)

Lead_clean = Lead %>% mutate(Group = ifelse(lead_grp == 1, "unexposed", "exposed"))
head(Lead_clean)
```

(b)<br>
```{r a2, message=FALSE, warning=FALSE, fig.align = "center"}
Unexposed = Lead_clean %>% filter(Group == "unexposed") %>% select(id, iqp)
boxplot(
  Unexposed$iqp,
  main = "IQP (Unexposed)",
  xlab = "IQP",
  col = "lightblue"
)

Q1_un = quantile(Unexposed$iqp, 0.25) 
Q3_un = quantile(Unexposed$iqp, 0.75) 
IQR_value_un = Q3_un - Q1_un                 

lower_bound_un = Q1_un - 1.5 * IQR_value_un
upper_bound_un = Q3_un + 1.5 * IQR_value_un

outliers_un = Unexposed %>% 
  filter(iqp < lower_bound_un | iqp > upper_bound_un)

print(as.numeric(outliers_un$id))
```
The outliers ids are: 117,135,139

(c)<br>
```{r a3, message=FALSE, warning=FALSE, fig.align = "center"}
Exposed = Lead_clean %>% filter(Group == "exposed") %>% select(id, iqp)
boxplot(
  Exposed$iqp,
  main = "IQP (Exposed)",
  xlab = "IQP",
  col = "lightblue"
)

Q1_e = quantile(Exposed$iqp, 0.25) 
Q3_e = quantile(Exposed$iqp, 0.75) 
IQR_value_e = Q3_e - Q1_e                 

lower_bound_e = Q1_e - 1.5 * IQR_value_e
upper_bound_e = Q3_un + 1.5 * IQR_value_e

outliers_e = Exposed %>% 
  filter(iqp < lower_bound_e | iqp > upper_bound_e)

print(as.numeric(outliers_e$id))
```
The outlier id is 314.

For those outliers, we'll perform the test both with and without the outliers. If the results are similiar, then the outliers can be included. If the results varies, then the outliers should be omitted.
