---
title: "HW6_1700"
author: "Yue Zhang"
date: "2024-11-04"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    keep_tex: true
  bookdown::html_document2:
    df_print: paged
    toc: true
geometry: margin=2.54cm
fontsize: 12pt
mainfont: Times New Roman
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
knitr::opts_knit$set(root.dir = "E:/Biostat/Biostatistics/PHL_1700/")
```

```{r packages, include = FALSE}
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(readxl)
library(haven)
library(stats)
library(pwr)

mytheme = theme_clean(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "right",
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
    ),
    plot.title = element_text(size = 12)
  )
theme_set(mytheme)
```

#10.14

|                        | Used OC Oracon | Did Not Use OC Oracon | Total |
|------------------------|----------------|-----------------------|-------|
| **Endometrial Cancer** | 6              | 111                   | 117   |
| **Control**            | 8              | 387                   | 395   |
| **Total**              | 14             | 498                   | 512   |

#10.20
```{r 2, message=FALSE, warning=FALSE}
phat_ear = 4/76
phat_ear
CI_L = p_hat-1.96*sqrt(p_hat*(1-p_hat)/76)
CI_L
CI_U = p_hat+1.96*sqrt(p_hat*(1-p_hat)/76)
CI_U
```
The point estimation is 0.05, and the 95% CI for the prevalence of otorrhea for the ear drop group is (0.002, 0.103)

#10.21
Two sample Z test can be used to compare the prevalence of otorrhea for the ear drop group to the observation group.
Null hypothesis: there's no difference in the prevalence of otorrhea between the ear drop group and the observation group.
Alternative hypothesis: there's difference in the prevalence of otorrhea between the ear drop group and the observation group.

#10.22
```{r 3, message=FALSE, warning=FALSE}
phat_obs = 41/75
p_pooled = (41+4)/(76+75)
se = sqrt(p_pooled*(1-p_pooled)*(1/76+1/75))
z = (phat_ear-phat_obs)/se
p_value = 2*(1-pnorm(abs(z)))
```
The p-value < 0.001, which means that we have enough evidence to reject the null hypothesis. Thus, the prevalence of otorrhea for the ear drop group is different than the prevalence of otorrhea for the observation group.

#10.38
Null hypothesis: the proportion of hens whose pancreatic secretions increased was equal across saline and secretin.
Alternative hypothesis: the proportion of hens whose pancreatic secretion increased was not equal across aline and secretin.
We should perform the Fisher's exact test.
```{r 4, message=FALSE, warning=FALSE}
Hormone = read_dta("/Biostat/Biostatistics/PHL_1700/Data/Raw/HORMONE-1.DAT.dta")
write.csv(Hormone, "Hormone.csv", row.names = FALSE)


#Take the first measure of each hen for each hormone administered
Hormone2 = Hormone %>% group_by(id, Hormone) %>% slice(1) %>% ungroup()

#Since we're only comparing the percentage of hens whose pancreatic secretions were different when comparing saline to secretin, we'll only select Hormone types equal to saline and secretin
Hormone2 = Hormone2 %>% filter(Hormone %in% c(1, 4))
head(Hormone2)

#Define a binary indicator for pancreatic secretion increase
Hormone2$Panph_Increase = ifelse(Hormone2$Pansecpt - Hormone2$Pansecpr>0, "Increase", "No Increase")

#Create a contingency table
table_hen = table(Hormone2$Hormone, Hormone2$Panph_Increase)
rownames(table_hen)= c("Saline", "Secretin")
table_hen = addmargins(table_hen, FUN = sum)
table_hen

#Calculate expected cells
c11 = 8*17/40
c11
c12 = 17*32/40
c12
c21 = 23*8/40
c21
c22 = 32*23/40
c22

#Since we have two expected cells less than 5, we could only use the Fisher's exact test.
fisher_hen = fisher.test(table_hen, alternative = "two.sided")
fisher_hen
```
Since the p-value is 0.893 which is larger than 0.05, we fail to reject the null hypothesis and conclude there is not enough evidence to support that the proportion of hens whose pancreatic secretions increased was equal across saline and secretin. I did the coding in R, and I've also tried in Stata, my two-sided p-value is 0.428 in Stata, which is still larger than 0.05.

#10.104
```{r 5, message=FALSE, warning=FALSE}
p1 = 200/10^5
p2 = p1*1.2
alpha = 0.05
power = 0.8
sample_size = power.prop.test(p1 = p1, p2 = p2, sig.level = alpha, power = power, alternative = "two.sided")
sample_size
```
There shall be 215368 women in each group.

#10.105
```{r 6, message=FALSE, warning=FALSE}
n = 20000
power2 = power.prop.test(p1 = p1, p2 = p2, sig.level = alpha, n = n, alternative = "two.sided")
power2
```
The power is 0.134, while in Stata, the estimated power = 0.136.

#10.125
We can use the McNemar's Test
```{r 7, message=FALSE, warning=FALSE}
standard_correct = c(82, 5)
standard_incorrect = c(10,3)
screening_test = data.frame(standard_correct, standard_incorrect)
row.names(screening_test) = c("new_correct", "new_incorrect")
screening_test= as.matrix(screening_test)
screening_test = addmargins(screening_test, FUN = sum)

#We have 15 discordant pairs which is fewer than 20, therefore, we cannot use the large sample method.
mcnemar_result = mcnemar.test(screening_test, correct = FALSE)
mcnemar_result
```
The p-value is 0.3921,in Stata, the p-value is 0.3018.
