---
title: "HW3_1830"
author: "Yue Zhang"
date: "2024-10-11"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r packages, include = FALSE}
setwd("E:/Biostat/Biostatistics/PH 1830/")
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(readxl)
library(lmtest)

```

# 3.3
(a)
P(Y=1) = 0.00109*X + 0.00255

(b)
```{r 1b, message=FALSE, warning=FALSE}
#(i)
#For X = 0
y1 = 0.00109*0 + 0.00255
y1
#For X = 7
y2 = 0.00109*7 + 0.00255
y2

#(ii)
RelativeRisk = y2/y1
RelativeRisk
```

(c)
```{r 1c, message=FALSE, warning=FALSE}
X = c(0, 1, 2, 3, 4)
Y1 = c(48, 38, 5, 1, 1)
N = c(17114, 14502, 793, 127, 38)
Table1 = data.frame(X, Y1, N)
fit1 = glm(Y1/N ~ X, family = quasi(link = "identity", variance = "mu(1-mu)"), weights = N, data = Table1)
summary(fit1)

#Based on the re-fit model: P(Y=1) = 0.00050*X + 0.00260 
#For X = 0
y3 = 0.00050*0+0.00260
y3
#For X = 7
y4 = 0.00050*7 + 0.00260
y4

RelativeRisk2 = y4/y3
RelativeRisk2

#Since the relative risk changes from 3.99 to 2.35 which is quite dramatic, this suggests that the results are sensitive to the choice of scores.
```

(d)
```{r 1d, message=FALSE, warning=FALSE}
#Linear model without the observation
X2 = c(0, 0.5, 1.5, 4.0)
Y2 = c(48, 38, 5, 1)
N2 = c(17114, 14502, 793, 127)
Table2 = data.frame(X2, Y2, N2)
fit2 = glm(Y2/N2 ~ X2, family = quasi(link = "identity", variance = "mu(1-mu)"), weights = N2, data = Table2)
summary(fit2)

#Based on the above model: P(Y=1) = 0.00082*X + 0.00260
#For X = 0
y5 = 0.00082*0+0.00260
y5
#For X = 7
y6 = 0.00082*7 + 0.00260
y6

RelativeRisk3 = y6/y5
RelativeRisk3

#Since we've calculated that the relative risk for linear model with the observation is 3.99, the relative risk decreases to 3.21 when omitting the observation. Therefore, the results are slightly sensitive to this single malformation.
```

#3.6
(a)
log[P(Y=1)/(1-P(Y=1))]=-0.5901*X + 3.1870
The negative coefficient for X indicates that as political ideology becomes more conservative, the likelihood of being a Democrat decreases.

(b)
```{r 2b, message=FALSE, warning=FALSE}
CI_L = -0.5901 - 1.96*0.1564
CI_L
CI_U = -0.5901 + 1.96*0.1564
CI_U
#The 95% Wald CI is (-0.896644, -0.283556) while the profile likelihood interval is (-0.91587 -0.29832). The profile likelihood interval is slightly wider than the 95% Wald CI as it accounts for the shape of the likelihood function, offering a more accurate estimate.
```

(c)
Z = -3.772
P_value = 0.000162
Since the p-value is less than alpha(0.05), we can reject the null hypothesis and conclude that the political ideology has a significant effect on the probability of being a Democrat.

(d)
```{r 2d, message=FALSE, warning=FALSE}
Democrat = c(5, 18, 19, 25, 7, 7, 2)
N3 = c(6, 21, 20, 36, 17, 18, 3)
X3 = c(1, 2, 3, 4, 5, 6, 7)
fit3 = glm(Democrat/N3 ~ X3, family = binomial(link = "logit"), weights = N3)
summary(fit3)

fit4 = glm(Democrat/N3 ~ 1, family = binomial(link = "logit"), weights = N3)
summary(fit4)
lrtest(fit3, fit4)

#The test statistic(chisq) is 17.009, and p-value is 3.72e-0.5 which is much smaller than alpha(0.05). Therefore, we can reject the null hypothesis and state that X3 has a significant effect on the probability of being a Democrat.
```

(e)
The number of Fisher Scoring iterations is 4 indicates that the model converged efficiently, as Fisher scoring is an iterative process used to estimate the parameters of the model.

#3.8
```{r 3, message=FALSE, warning=FALSE}
Snoring = c(0, 2, 4, 5)
HeartDisease_Y = c(24, 35, 21, 30)
HeartDisease_N = c(1355, 603, 192, 224)
Table3 = data.frame(Snoring, HeartDisease_Y, HeartDisease_N)
Table3$N4 = Table3$HeartDisease_Y + Table3$HeartDisease_N
fit5 = glm(HeartDisease_Y/N4 ~ Snoring, family = binomial(link = "logit"), weights = N4, data = Table3) 
summary(fit5)
confint(fit5)
```

#3.13
(a)
```{r 4a, message=FALSE, warning=FALSE}
Crab = read.table("http://www.stat.ufl.edu/~aa/cat/data/Crabs.dat", header = TRUE)
fit6 = glm(sat ~ weight, family = poisson(link = "log"), data = Crab)
summary(fit6)

#Based on above, the prediction equation is: mu_hat = exp(0.58930*X - 0.42841)
#For x = 2.44kg,
MeanResponse = exp(0.58930*2.44-0.42841)
MeanResponse
```

(b)
```{r 4b, message=FALSE, warning=FALSE}
exp(0.58930)
#The beta hat means that for every 1kg increase in crab weight, the expected number of satellites increases by approximately 80.3%.

CI_L2 = 0.58930 - 1.96*0.06502
CI_L2
CI_U2 = 0.58930 + 1.96*0.06502
CI_U2
#The 95% CI is (0.461861, 0.716739). 

exp(0.461861)
exp(0.716739)
#The 95% CI for multiplicative effect is (1.58703, 2.04775)
```

(c)
```{r 4c, message=FALSE, warning=FALSE}
#Wald test
#Null hypothesis: beta = 0, Alternative hypothesis: beta is not equal to 0.
Z_squared = (0.58930/0.06502)^2
Z_squared

#Likelihood Ratio test
fit7 = glm(sat ~ 1, family = poisson(link = "log"), data = Crab)
summary(fit7)
lrtest(fit6, fit7)
#Since the Likelihood ratio test statistic is 71.925 and df = 1, and the p-value is far less than 0.05 which means that we can reject the null hypothesis.

```

