---
title: "HW6_1830"
author: "Yue Zhang"
date: "2024-12-01"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r packages, include = FALSE}
setwd("E:/Biostat/Biostatistics/")
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(readxl)
library(lmtest)
library(mfx)
library(pROC)
library(VGAM)

mytheme = theme_clean(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "right",
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
    ),
    plot.title = element_text(size = 12)
  )
theme_set(mytheme)
```

#1
```{r 1, message=FALSE, warning=FALSE}
afterlife = read.table("http://www.stat.ufl.edu/~aa/cat/data/Afterlife.dat", header = TRUE)
afterlife_model = vglm(cbind(yes, undecided, no) ~ gender + race, family = multinomial, data = afterlife)
summary(afterlife_model)

#(a)
exp(-0.105)

#(b)
exp(-0.429-(-0.105))
```
(a) The odds ratio for undecided and no pair is 0.9 which means that men are 0.9 times likely to response undecided rather than no than females.
(b) The odds ratio for yes and undecided pair is 1.311 which means that men are 1.311 times likely to response yes rather than undecided than females.

#2
```{r 2, message=FALSE, warning=FALSE}
happy = data.frame(
  income = 1:3,
  y1 = c(6, 6, 6),
  y2 = c(43, 113, 57),
  y3 = c(75, 178, 117)
)

happy_model = vglm(cbind(y1, y2, y3) ~ income, data = happy, family = cumulative(parallel = TRUE))
summary(happy_model)

happy_model2 = vglm(cbind(y1, y2, y3) ~ income, data = happy, family = cumulative(parallel = FALSE))
summary(happy_model2)

lrtest(happy_model, happy_model2)
```
The p-value for the likelihood ratio test is 0.788 which is larger than 0.05. Therefore, we cannot reject the null hypothesis. This indicates that the proportional odds assumption holds and the simpler proportional odds model is appropriate for analyzing the effect of income on happiness. Thus, we can fit an ordinal regression model here.

#3
```{r 3, message=FALSE, warning=FALSE}
alligator = read.csv("E:/Biostat/Biostatistics/PH 1830/alligator.csv")
alligator_model = vglm(food ~ length, family = multinomial, data = alligator)
summary(alligator_model)

1 - exp(-0.110)
1 - exp(-2.465)
exp(2.355)

```
The ML prediction equations are: $logit(\frac{\hat\pi_1}{\hat\pi_3}) = 1.618 - 0.110*x$, $logit(\frac{\hat\pi_2}{\hat\pi_3}) = 5.698 - 2.465*x$, $logit(\frac{\hat\pi_1}{\hat\pi_2}) = -4.080 + 2.355*x$
For one unit increases in length, the odds of choosing fish rather than other decrease by 10.4%.
For one unit increases in length, the odds of choosing invertebrate rather than other decrease by 91.5%.
For one unit increases in length, the odds of choosing fish is 10.538 times of choosing invertebrate.
